---
title: "Homework #1"
author: "Suruchi Ahuja, Abbas Rizvi, Hoi Lam Tai, Jingchen Zhang"
date: "February 15, 2016"
output: pdf_document
---

# Problem 1
The dataset from Lec04.R was loaded. The following code was provided to us by Dr. Gaile.
```{r, message=FALSE}
require(knitr)
require(GEOquery)
load("gse19439.RData")
#so I am just going to load it now
load(file="gse19439.RData")
gse19439
```


Dr. Gaile creates a factor object `FTB` for the groups (control, latent TB, and positive TB) that corresponds to the samples (in the same order).
```{r}
# make a factor object for Control, latent TB and Positive TB
tmp <- as.character(pData(phenoData(gse19439))[,1])
J <- length(tmp) # J=number of samples
TBgroup <- rep("",J)
for(j in 1:J) TBgroup[j]=substring(tmp[j],1,3)
# make a factor for TBgroup
FTB <- factor(TBgroup,levels=c("CON","LTB","PTB"))
X <- exprs(gse19439)
# do a quick kruskal-wallis scan
myKrusk <- function(i){
        cat(i,"...",fill=F)
        kruskal.test(x=X[i,], g=FTB)$p.value
}

```

The p-values that were calculated from the Kruskal-Wallis test were recorded into `myPvals.RData`. The best 4 p-values (`myPvals`) were subsequently sorted in ascending order (the lowest value at the beginning of the vector). This order was assigned to groups in STA 525.

```{r}
load("myPvals.RData")
GroupLabels <- c("Group I","Group II","Group III","Group IV")
# pick the best 4 p-values and assign them to the students.
best4 <- order(myPvals)[1:4]
# print out list of best 4
print(best4)

for(i in 1:length(GroupLabels)){
        cat("Group Label:", GroupLabels[i], "\t\t row assignment:", best4[i], fill=T)
}
myrow <- gse19439[10685,]
dim(myrow)
```

Since we are group II, our row assignment was determined to be 10685 (as shown above). We were asked whether or not the assigned row corresponded to a specific gene and if so, what is known about the gene. Since each row does correspond to a gene, we subsetted `ExpressionSet` just to our assigned row and pulled out the `featureData`. 

```{r}
featureNames(myrow)
gene.info <- pData(featureData(myrow))
gene.info
```


# Problem 2
## 2.1 Boxplot
The data was log2 transformed for scability.
```{r}
sampleNames(myrow)
boxplot(log2(X[10685,sampleNames(myrow)[FTB == 'CON']]),
        log2(X[10685,sampleNames(myrow)[FTB == 'LTB']]),
        log2(X[10685,sampleNames(myrow)[FTB == 'PTB']]),
        main = "Boxplot of Row Data as Function of TB Phenotype",
        names = c("CON", "LTB", "PTB"))
#boxplots disguise multimodal data 
```

## 2.2 Violin Plot
```{r, message=FALSE}
require(vioplot)
vioplot(log2(X[10685,sampleNames(myrow)[FTB == 'CON']]),
        log2(X[10685,sampleNames(myrow)[FTB == 'LTB']]),
        log2(X[10685,sampleNames(myrow)[FTB == 'PTB']]),
        names = c("CON", "LTB", "PTB"))
title("Violin Plot of Row Data as Function of TB Phenotype")
```

## 2.3 Bean Plot
```{r, message=FALSE}
require(beanplot)
beanplot(log2(X[10685,sampleNames(myrow)[FTB == 'CON']]),
        log2(X[10685,sampleNames(myrow)[FTB == 'LTB']]),
        log2(X[10685,sampleNames(myrow)[FTB == 'PTB']]),
        names = c("CON", "LTB", "PTB"))
title("Bean Plot of Row Data as Function of TB Phenotype")
```


# Problem 3
## 3.1 Heatmaps
The best 20 student features were clustered and plotted in a heatmap using `heatmap.2`. The clustering method worked reasonably well in determined what group the samples belonged to (see column labels of heatmap). 
```{r, message=FALSE}
require(gplots)
best20 <- order(myPvals)[1:20]
x.best <- X[best20,]
heatmap.2(X[best20,], main="20 student features", 
          scale = "row", trace="none", labCol = FTB)
```



# Problem 4
## 4.1 Pre-processing
We removed NAs and zeroes.
```{r}
spikeDF <- read.table(file="AffyProbeSpikeValues.csv",sep="\t")
levels(spikeDF[,2])
summary(spikeDF[,2])

#grab Spike fold changes for all entries as numeric ...
#...so anything that was not a number is now an NA
SpikeFC <- as.numeric(levels(spikeDF[,2])[spikeDF[,2]])

#grab IDs
names(SpikeFC) <- spikeDF$V1

#remove NAs
nonZeroDX <- which(!is.na(SpikeFC) & (SpikeFC != 0)) 
spikeFC.clean <- SpikeFC[nonZeroDX]

#check how many genes are left in dataset
length(spikeFC.clean)
```

## 4.2 Normalization of 8 routes with expresso() and subset into ExpressionSet
```{r, eval=FALSE}
require(affy)
bgcorrect.mtd <- c("rma", "mas", "rma", "mas",
                   "rma", "none", "mas", "mas")
normalized.mtd <- c("constant", "quantiles", "quantiles", "loess",
                    "loess", "constant", "qspline", "qspline")
pmcorrect.mtd <- c("pmonly", "pmonly", "subtractmm", "mas",
                   "mas", "pmonly", "subtractmm", "mas")
summary.mtd <- c("mas", "mas", "avgdiff", "medianpolish", 
                 "medianpolish", "avgdiff", "mas", "mas")

route.expsets <- list()
for (i in 1:length(bgcorrect.mtd)){
        routes <- expresso(affydata, 
                           bgcorrect.method = bgcorrect.mtd[i], 
                           normalize.method = normalized.mtd[i],
                           pmcorrect.method = pmcorrect.mtd[i], 
                           summary.method = summary.mtd[i])
        route.expsets[[i]] <- exprs(routes)[nonZeroDX,]
}
```

## 4.3 Multiple hypothesis testing
```{r,eval=FALSE}
#create labels for multitesting class labels -- 18 samples, 9 control, 9 experimental
labels <- factor(c(rep(0, 9), rep(1, 9))) #0 is control, 1 is experimental

#source("http://bioconductor.org/biocLite.R")
#biocLite("multtest")				
library(multtest)
stats <- list()
for (i in 1:length(route.expsets)){
        #conduct multiple t test for 10000 permutations  
        testing.routes <- mt.maxT(route.expsets[[i]],
                                  classlabel = labels, B = 10000) 
        stats[[i]] <- testing.routes$teststat[order(testing.routes$index)]
}

nrow <- nrow(route.expsets[[1]])
myresponse <- rep(NA, nrow)
#if the spike value is 1 ... assign as 0s in matrix ... 1s are the control
myresponse[which(spikeFC.clean == 1)] = 1 
#if spike value is not 1 ... assign as 1s in the matrix ... 0s are exp.
myresponse[which(spikeFC.clean != 1)] = 0 
```

## 4.4 ROC Curves
```{r, eval=FALSE}
#apply roc function on input 
roc.fnct <- function(x){roc(response = myresponse,
                            predictor = abs(x), plot=TRUE, print.auc=TRUE)}
#apply roc function on all the ordered test statistics in the list stats
roc <- lapply(stats, roc.fnct) 
```

```{r, message=FALSE}
load("question4.RData")
rainbow <-  palette(rainbow(length(route.expsets)))
plot(roc[[1]], main = "ROC curves for different normalization routes using expresso()", 
     col=rainbow[1])
legendText <- c()
for(i in 1:length(route.expsets)){
        plot(roc[[i]], add=TRUE, col=rainbow[i])
        legendText[i] <- paste(bgcorrect.mtd[i],
                               "/",normalized.mtd[i],"/", 
                               pmcorrect.mtd[i],
                               "/",summary.mtd[i],
                               "   AUC: ",
                               round(as.numeric(roc[[i]]$auc),3), sep="")
}
legend("bottomright",
       legendText,
       lty=c(rep(1,length(route.expsets))), 				
       lwd=c(rep(3,length(route.expsets))),
       col=rainbow)
```
